Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-10 E-Hypergraphs-LICS24.tex
conclusion

We have described a categorical semantics for e-graphs over algebraic theories in terms of Cartesian categories enriched over a semilattice, and defined a natural generalization to account for monoidal theories in terms of SMCs,  similarly enriched.  Having described a translation of e-graphs into our formalism, we developed a combinatorial representation of morphisms in our free category which factor out the structural SMC equations, but remain sensitive to the complexity-relevant equations induced by the enrichment.  The design of such a representation and its corresponding  rewrite theory proved delicate, and---together with its proof of completeness---constitute the main technical contributions of the paper.  Finally, we considered an analogue of e-rewriting in our setting, pushing towards the practical considerations necessary for an implementation of our ideas.  Although we believe our work forms a solid basis for a formal mathematical theory of (acyclic) e-graphs,  there are a number of extensions which we intend to investigate in the future.

\subsection{Future Work and Applications}

We have already alluded to the fact we deal with only \textit{acyclic} e-graphs in this paper.  Nevertheless,  within this setting we present a generalization from Cartesian to (symmetric) monoidal theories,  giving rise to a host of potential new domains of application.  Further,  the string diagrammatic and graph rewriting framework we set up is ameenable to extension, including lifting the acyclicity restriction.

\begin{figure}
\[
	\scalebox{0.75}{
	\tikzfig{conclusion/example-e-graphs}
	}
\]
\[
	\scalebox{0.75}{
	\tikzfig{conclusion/example-traced}
	}
\]
\caption{Example applications of e-hypergraphs.  Top left, the ZX-calculus.  Top right, the $\lambda$-calculus as represented by string diagrams for Cartesian closed categories.  Below,  an extension of Figure \ref{fig:} by another rewrite step, $1 * x \to x$,  introducing a cycle modelled by traced structure. }
\end{figure}\label{fig:applications}

Symmetric monoidal theories have received significant attention over recent years: the ZX-calculus can be used to characterize quantum processes, with applications to optimization of quantum circuits \cite{coecke_interacting_2011,ZX} ; similar theories also characterize signal flow diagrams 
\cite{baldan_categorical_2014}, digital circuits 
\cite{ghica_operational_2021}, as well as having application to computational linguistics \cite{wazni_quantum_2022,coecke_lambek_2013}.  In future work, we hope to apply our generalization of e-graphs to optimization problems in these domains (e.g. Figure \ref{fig:applications})

The theories described above often have more structure than simply being symmetric monoidal: for example, they may contain comonoids, Frobenius algebras or traced structure (as in the case of e-graphs with cycles),  and this structure can be exploited to give more natural (in the sense of factoring out more equations) combinatorial representations.  In the above cases, such representations and their DPO-rewriting theory is well-studied 
\cite{ghica_rewriting_2023} % traced comonoid
[cite, cite, cite].  We consider that our development of hierarchical "e-box" structure within string diagrams and their associated (e-)hypergraphs ( categorically,  semilattice enrichment) is essentially orthogonal to these developments, and so ought to be easily combined to give notions of e-hypergraphs specifically suited to the optimization of these particular monoidal theories.  A particularly interesting extension that we intend to consider is the combination of our "e-structure" with (Cartesian) \textit{closed} categories, allowing the natural combination of e-graph optimization techniques with the $\lambda$-calculus.  The corresponding combinatorial representation would build on the recent work \cite{ghica_hierarchical_2023, alvarez-picallo_rewriting_2022, alvarez-picallo_functorial_2021}.  String diagrams for these various examples are included in Figure \ref{fig:applications}.  There is also the question of developing a corresponding term calculus \cite{Heijltjes:FMCII}. % hierarchical

>>>>

We have seen how e-graphs over algebraic theories can be expressed in terms of Cartesian categories enriched over a semilattice. 
In fact, e-graphs are an instance of a more general construction that can account for monoidal theories in terms of SMCs,  similarly enriched.  
Having described a translation of e-graphs into this categorical framework, a combinatorial representation of morphisms in the free category is also provided. 
As expected, this representation factors out the structural SMC equations, while remaining sensitive to the complexity-relevant equations induced by the enrichment.  The design of such a representation and its corresponding rewrite theory proved to be delicate, and, together with its proof of completeness, constitute the main technical contributions of the paper.  Finally, we considered an analogue of e-rewriting in this setting, required in the developing potential implementations.  In addition to the current mathematical treatment of (acyclic) e-graphs,  there are a number of extensions which we intend to investigate in the future.

Some of the extensions that can be considered are: support for `functorial boxes'~\cite{mellies} (in addition to e-boxes), which have a variety of applications, including the Cartesian-closure required for the implementation of functional programming languages~\cite{GhicaZanasiTutorial}, as sketched in Fig. \ref{fig:app1}(a); the `spiders' used by the ZX-calculus to model quantum circuits~\cite{coecke_interacting_2011,ZX}, as sketched in Fig. \ref{fig:app1}(b); or trace~\cite{Joyal, Hasegawa} which can be used to express feedback in categorical models of digital circuits~\cite{GhicaJung, GhicaKaye} and, indeed in conventional e-graphs to encode infinite equivalence classes, as sketched in Fig.~\ref{fig:app2}. All these application domains stand to benefit from the use of optimisation techniques, which in turn can take advantage of the appropriate e-graph technique.

\begin{figure}
\[
	\scalebox{0.75}{
	\tikzfig{conclusion/example-e-graphs}
	}
\]
\caption{E-graphs for functorial boxes and the ZX-calculus
DAN: THE LINE BETWEEN THE APPLICATION NODE AND THE BOX IS TOO LONG}
\label{fig:app1} 
\end{figure}
\begin{figure}
\[
	\scalebox{0.75}{
	\tikzfig{conclusion/example-traced}
	}
\]
\label{fig:app2}
\caption{E-graphs with cycles and e-hypergraphs with trace}
\end{figure}

The theories described above often have more structure than simply being symmetric monoidal: for example, they may contain comonoids, Frobenius algebras or traced structure (as in the case of e-graphs with cycles),  and this structure can be exploited to give more natural (in the sense of factoring out more equations) combinatorial representations.  In the above cases, such representations and their DPO-rewriting theory is well-studied 
\cite{ghica_rewriting_2023,cite,cite,cite}. % traced comonoid

The hierarchical ``e-box'' structure within string diagrams and their associated (e-)hypergraphs (categorically, the semilattice enrichment) is essentially orthogonal to these developments. Combining the two, to give notions of e-hypergraphs specifically suited to these particular monoidal theories should be conceptually straightforward. However, as we have seen in this paper, the technical details can be delicate. 
Finally, a term calculus (e-terms) can be reverse engineered from the string-diagram notation~\cite{https://doi.org/10.4204/EPTCS.266.4}. Although term calculi are less important in the algorithmics of optimisation, they still play a role in specification and in aiding readability.
